---
title: "Intraspecific Trait Variation"
author: "Orlando Acevedo-Charry, Brynn M. Fricke, Mia Keriazes"
date: "2025-02-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Intro and packages

bhla bhla bhla

```{r call packages}
# data manipulation and visualization
library(tidyverse)

# correlation plot
library(corrplot)

# Functional diversity 
library(funspace); library(TPD)
```

## Calling data

### Measurements in the museum

```{r data measured}
data_specimens <- read_csv("flmnh_data/Museum Data Updated.csv") |>
  mutate(species_updated = paste(Genus, Species, sep = " "),
         body.mass = as.numeric(Mass))

species <- unique(data_specimens$species_updated)

```

### AVONET raw data

We use the raw data (`AVONET_Raw_Data` sheet in the Excel file) in AVONET ([Tobias et al. Ecol Letters 2022](https://onlinelibrary.wiley.com/doi/10.1111/ele.13898)), to estimate a mean and SD of trait per species. Unfortunately, raw data of AVONET does not include the variation of Body mass, they only included the mean value in their summary (`AVONET2_eBird` sheet in the Excel file). We will pull the other sources of functional traits and estimate mean and SD of this trait afterwards. We also include the taxonomy information in AVONET (eBird).

```{r AVONET raw data}
avonet <- read_csv("FunctionalTraits/AVONET_Raw_Data.csv") |> 
  mutate(species_updated = ifelse(eBird.species.group == "Chlorophonia musica flavifrons",
                                  "Chlorophonia flavifrons",
                           ifelse(eBird.species.group == "Chlorophonia musica musica",
                                  "Chlorophonia musica",
                           ifelse(eBird.species.group == "Chlorophonia musica sclateri",
                                  "Chlorophonia sclateri",
                           Species2_eBird)))) |>
  dplyr::select(species_updated, Source, Specimen.number,
                Beak.Length_Culmen, Beak.Width, Beak.Depth,
                Tarsus.Length, `Hand-wing.Index`, Tail.Length) |>
  rename(HWI = `Hand-wing.Index`) |>
  filter(species_updated %in% species)

```
 
This raw data does not have Mass values. We can extracted from the summarized `AVONET2_eBird` dataset.

```{r AVONET2_eBird}
avonet2_eBird <- read_csv("FunctionalTraits/AVONET2_eBird.csv") |>
  dplyr::select(Species2,
                Mass, `Hand-Wing.Index`, Tarsus.Length, Tail.Length, 
                Beak.Length_Culmen, Beak.Width, Beak.Depth) |>
  rename(body.mass = Mass) |>
  rename(HWI = `Hand-Wing.Index`)

avonet_raw <- avonet |>
  left_join(avonet2_eBird |> dplyr::select(Species2, body.mass), 
            join_by("species_updated" == "Species2"))
```

Finally, we added the archipelago group (manually) to these 383 specimens

```{r island or mainland for avonet}
avonet_islands <- read_csv("flmnh_data/AVONET_b_m_specimens_to_review.csv") |> 
  left_join(avonet_raw)
```

### Combine data

The NA data will be replaced by the mean reported in AVONET 

```{r generate unified traits data}
names(data_specimens)
names(avonet_islands)

traits_specimens <- data_specimens |> 
  dplyr::select(Archipelago_group, species_updated,
                body.mass, HWI, Tarsus.Length, Tail.Length, 
                Beak.Length_Culmen, Beak.Width, Beak.Depth)

traits_avonet <- avonet_islands |>
  dplyr::select(Archipelago_group, species_updated,
                body.mass, HWI, Tarsus.Length, Tail.Length, 
                Beak.Length_Culmen, Beak.Width, Beak.Depth)

data_traits <- traits_specimens |>
  rbind(traits_avonet) |>
  group_by(species_updated) |>
  mutate(body.mass = ifelse(is.na(body.mass), avonet2_eBird$body.mass,
                            body.mass),
         HWI = ifelse(is.na(HWI), avonet2_eBird$HWI,
                            HWI),
         Tarsus.Length = ifelse(is.na(Tarsus.Length), avonet2_eBird$Tarsus.Length,
                            Tarsus.Length),
         Tail.Length = ifelse(is.na(Tail.Length), avonet2_eBird$Tail.Length,
                            Tail.Length),
         Beak.Length_Culmen = ifelse(is.na(Beak.Length_Culmen), avonet2_eBird$Beak.Length_Culmen,
                            Beak.Length_Culmen),
         Beak.Width = ifelse(is.na(Beak.Width), avonet2_eBird$Beak.Width,
                            Beak.Width),
         Beak.Depth = ifelse(is.na(Beak.Depth), avonet2_eBird$Beak.Depth,
                            Beak.Depth))
```

Correct some names and add Id for each data, grouping by species

```{r adjust names et al}
colnames(data_traits) <- c("Archipelago", "Species",
                           "mass", "h.w.i", "tars.l", "tail.l",
                           "beak.l", "beak.w", "beak.d")

data_traits <- data_traits |>
  group_by(Species, Archipelago) |>
  mutate(Sequence = row_number()) |>
  ungroup()

# a function for extract strings
id_specimen <- function(archipelago, species) {
    archipelago_part <- str_extract_all(archipelago, 
                                        "\\b\\w{2}") |>
      unlist() |>
      paste0(collapse = "")
    
    species_part <- str_extract_all(species, 
                                    "\\b\\w{3}") |>
      unlist() |>
      paste0(collapse = "")
    
    paste0(species_part, "_", archipelago_part)
}

data_traits <- data_traits |>
  mutate(Group = mapply(id_specimen, Archipelago, Species),
         ID = paste0(Group, "_", Sequence))
```

## Functional Trait Space - PCA based 

### Chlorophonia-Euphonia

```{r eup-scale and names}
eup_traits <- data_traits |>
  filter(Species != "Coereba flaveola")

eup_traits_scl <- as.data.frame(scale(eup_traits[c(3:9)]))

corrplot(round(cor(eup_traits_scl),2), type="upper", order="hclust", 
         tl.col="black", tl.srt=45)

rownames(eup_traits_scl) <- eup_traits$ID

# Calculating the dimensionality of a functional space based on PCA
funspaceDim(eup_traits_scl)
```

Run PCA of the scaled traits and building the functional trait spaces

```{r eup-Run PCA}
pca.traits.eup <- princomp(eup_traits_scl, cor = T)

plot(pca.traits.eup)
```

And we can do a functional trait space analysis grouping by species and archipelago

```{r eup-run trait space global and by group}
# trait space by groups
trait.space.pca.eu.global <- funspace(x = pca.traits.eup,
                               n_divisions = 200)
# trait space by Archipelago
trait.space.pca.eu.arch <- funspace(x = pca.traits.eup,
                               group.vec = eup_traits$Archipelago,
                               n_divisions = 200)

# trait space by group of species by archipelago
trait.space.pca.eu.group<- funspace(x = pca.traits.eup,
                               group.vec = eup_traits$Group,
                               n_divisions = 200)
```

We can explore the results 

```{r eup-explore trait space global}
summary(trait.space.pca.eu.global)
plot(x = trait.space.pca.eu.global, type = "global", 
     quant.plot = T, arrows = T, arrows.length = 0.75)
```

```{r eup-explore trait space archipelago}
summary(trait.space.pca.eu.arch)
plot(x = trait.space.pca.eu.arch, type = "groups", 
     quant.plot = T,globalContour = T, pnt = T, pnt.cex = 0.1)
```

```{r eup-explore trait space group}
summary(trait.space.pca.eu.group)
plot(x = trait.space.pca.eu.group, type = "groups", 
     quant.plot = T, globalContour = T, pnt = T, pnt.cex = 0.1)
```

### Coereba flaveola

```{r coe-scale and names}
coe_traits <- data_traits |>
  filter(Species == "Coereba flaveola")

coe_traits_scl <- as.data.frame(scale(coe_traits[c(3:9)]))

corrplot(round(cor(coe_traits_scl),2), type="upper", order="hclust", 
         tl.col="black", tl.srt=45)

rownames(coe_traits_scl) <- coe_traits$ID

# Calculating the dimensionality of a functional space based on PCA
funspaceDim(coe_traits_scl) 
```

Run PCA of the scaled traits and building the functional trait spaces

```{r coe-Run PCA}
pca.traits.coe <- princomp(coe_traits_scl, cor = T)

plot(pca.traits.coe)
```

And we can do a functional trait space analysis grouping by species and archipelago

```{r coe-run trait space global and by group}
# trait space by groups
trait.space.pca.coe.global <- funspace(x = pca.traits.coe,
                               n_divisions = 200)
# trait space by Archipelago
trait.space.pca.coe.arch <- funspace(x = pca.traits.coe,
                               group.vec = coe_traits$Archipelago,
                               n_divisions = 200)
```

We can explore the results 

```{r coe-explore trait space global}
summary(trait.space.pca.coe.global)
plot(x = trait.space.pca.coe.global, type = "global", 
     quant.plot = T, arrows = T, arrows.length = 0.75)
```

```{r coe-explore trait space archipelago}
summary(trait.space.pca.coe.arch)
plot(x = trait.space.pca.coe.arch, type = "groups", 
     quant.plot = T, globalContour = T, pnt = T, pnt.cex = 0.1)
```
